---
title: "combined_growth_analysis"
author: "Dan Stoebel"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(growthrates)
library(patchwork)
library(RColorBrewer)
library(infer)
theme_set(theme_classic())
```

```{r}
files <- c("../../data/growth_data/growth_rates_20230519.xlsx", "../../data/growth_data/growth_rates_20230524.xlsx", "../../data/growth_data/growth_rates_20230525.xlsx", "../../data/growth_data/growth_rates_20230531.xlsx", "../../data/growth_data/growth_rates_20230602.xlsx")
merged <- files %>% map_dfr(read_excel, .id="filename", na = "NA")

```

```{r}
all_growth_data <- merged %>%
  pivot_longer(cols = contains("25"), values_to = "OD") %>%
  separate(col = name,
           into = c("strain", "stress"),
           sep = "_") %>%
  group_by(filename) %>%
  mutate(elapsed_mins = (clock_time[[1]] %--% clock_time) / dminutes(1)) %>%
  ungroup() %>%
  select(trial = filename, elapsed_mins, strain, stress, OD) %>%
  mutate(
    strain = fct_recode(strain, `wild-type` = "2537", `rpoS` = "2545"),
    stress = fct_relevel(stress, "nostress", "salt"),
    stress = fct_recode(
      stress,
      `Entry into stationary phase` = "nostress",
      `High osmolarity` = "salt",
      `Low temp` = "cold"
    ),
    trial = paste("Replicate", trial)
  )
  
```

```{r}
# Some quick checks on data quality
# View(all_growth_data)
# all_growth_data %>% 
#   group_by(trial) %>% 
#   skim()
```


Plot all of the data
```{r}
all_growth_data %>% 
  ggplot(aes(x = elapsed_mins, y = OD, col = strain)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(se = FALSE, na.rm = TRUE) +
  scale_y_log10() +
  facet_grid(stress~trial, scales = "free", space = "free") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "time (mins)",
       y = expression(OD[600])) +
  theme(strip.text.x = element_blank()) +
    NULL

ggsave("../../outputs/supplemental/all_growth_curves.pdf", width = 10, height = 6)
```


Fit the Gomperz 3 param model to the cold data
```{r}
p <- c(y0 = 0.3, mumax = .001, K = 1, lambda = 100)
cold_gompertz3 <- all_growth_data %>% 
  filter(stress == "Low temp") %>% 
  all_growthmodels(OD ~ grow_gompertz3(time = elapsed_mins, parms) |stress + trial + strain, data = ., p = p)

cold_estimates_raw <- coef(cold_gompertz3)
```


Fit the Gomperz 3 param model to the salt data
```{r}
p <- c(y0 = 0.3, mumax = .01, K = 1, lambda = 100)
salt_gompertz3 <- all_growth_data %>% 
  filter(stress == "High osmolarity") %>% 
  all_growthmodels(OD ~ grow_gompertz3(time = elapsed_mins, parms) |stress + trial + strain, data = ., p = p)

salt_estimates_raw <- coef(salt_gompertz3)
```

Fit the logistic model to the no stress data. I use the logistic instead of the gomperz 3 param model because there is no lag time in this growth experiment. Visiually, the logistic fits better than the gomperz 2 param model.

```{r}
p <- c(y0 = 0.3, mumax = .02, K = 6)

nostress_logistic <- all_growth_data %>% 
  filter(stress == "Entry into stationary phase") %>% 
  all_growthmodels(OD ~ grow_logistic(time = elapsed_mins, parms) |stress + trial + strain, data = ., p = p)

nostress_estimates_raw <- coef(nostress_logistic)
```

# Analysis of growth params

```{r}
salt_estimates <- salt_estimates_raw %>% 
  as_tibble(rownames = "rowname") %>% 
  separate(col = rowname, into =c("stress", "replicate", "strain"), sep = ":") %>% 
  mutate(doubling_time = log(2)/mumax)
```

For salt, I'm interested in if there is a difference in lag (lambda) or in growth rate (mumax). I'm not interested in K (carrying capacity) because cells don't approach K over this experiment. 
```{r}
salt_params_summary <- salt_estimates %>% 
  group_by(strain) %>% 
  summarise(n = n(),
            mean_lambda = mean(lambda),
            se_lambda = sd(lambda)/n,
            mean_doubling_time = mean(doubling_time),
            se_doubling_time = sd(doubling_time)/n
            )

write_tsv(salt_params_summary, "../../outputs/growth_curves/salt_growth_params.tsv")
```


```{r}
set.seed(2023)

salt_diff_mean <- salt_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

salt_doubling_pval <- salt_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = salt_diff_mean, direction = "two-sided")

salt_diff_lag <- salt_estimates %>% 
  specify(lambda ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

salt_lambda_pval <- salt_estimates %>% 
  specify(lambda ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = salt_diff_lag, direction = "two-sided")

salt_doubling_pval
salt_lambda_pval
```




```{r}
cold_estimates <- cold_estimates_raw %>% 
  as_tibble(rownames = "rowname") %>% 
  separate(col = rowname, into =c("stress", "replicate", "strain"), sep = ":") %>% 
  mutate(doubling_time = log(2)/mumax)
```

For cold, I'm interested in if there is a difference in lag (lambda) or in growth rate (mumax). I'm not interested in K (carrying capicity) because cells don't approach K over this experiment. 
```{r}
cold_params_summary <- cold_estimates %>% 
  group_by(strain) %>% 
  summarise(n = n(),
            mean_lambda = mean(lambda),
            se_lambda = sd(lambda)/n,
            mean_doubling_time = mean(doubling_time),
            se_doubling_time = sd(doubling_time)/n
            )
write_tsv(cold_params_summary, "../../outputs/growth_curves/cold_growth_params.tsv")
```


```{r}
set.seed(2023)
cold_diff_mean <- cold_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

cold_doubling_pval <- cold_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = cold_diff_mean, direction = "two-sided")

cold_diff_lag <- cold_estimates %>% 
  specify(lambda ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

cold_lambda_pval <- cold_estimates %>% 
  specify(lambda ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = cold_diff_lag, direction = "two-sided")

cold_doubling_pval
cold_lambda_pval

```



## No stress condition

```{r}
nostress_estimates <- nostress_estimates_raw %>% 
  as_tibble(rownames = "rowname") %>% 
  separate(col = rowname, into =c("stress", "replicate", "strain"), sep = ":") %>% 
  mutate(doubling_time = log(2)/mumax)
```

For the no added stress, I'm interested in if there is a difference in growth rate (mumax) or carrying capacity (K). There is no lag-time in this model.
```{r}
nostress_params_summary <- nostress_estimates %>% 
  group_by(strain) %>% 
  summarise(n = n(),
            mean_K = mean(K),
            se_K = sd(K)/n,
            mean_doubling_time = mean(doubling_time),
            se_doubling_time = sd(doubling_time)/n
            )

write_tsv(nostress_params_summary, "../../outputs/growth_curves/nostress_growth_params.tsv")
```


```{r}
set.seed(2023)
nostress_diff_mean <- nostress_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

nostress_doubling_pval <- nostress_estimates %>% 
  specify(doubling_time ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = nostress_diff_mean, direction = "two-sided")

cold_diff_K <- nostress_estimates %>% 
  specify(K ~ strain) %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS"))

nostress_K_pval <- nostress_estimates %>% 
  specify(K ~ strain) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("wild-type", "rpoS")) %>% 
  get_p_value(obs_stat = cold_diff_K, direction = "two-sided")

nostress_doubling_pval
nostress_K_pval
```

FDR adjustment of all p-values
```{r}

p.adjust(c(nostress_doubling_pval, nostress_K_pval, salt_doubling_pval, salt_lambda_pval, cold_doubling_pval, cold_lambda_pval), method = "BH") %>% round(2)

```


Create a pdf of all individual model fits
```{r eval=FALSE, include=FALSE}
pdf(file = "../../outputs/supplemental/all_model_fits.pdf", width = 20, height = 10)
par(mfrow=c(2,5))
plot(nostress_logistic, log = "y")
plot(salt_gompertz3, log = "y")
plot(cold_gompertz3, log = "y")
dev.off()
```

